---
title: "week_11_wederstrandt_d"
author: "David Wederstrandt"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo=FALSE)
```



```{r include=FALSE, warning=FALSE, message=FALSE}
library(readr)
library(tidyverse)
```


```{r include=FALSE, warning=FALSE, message=FALSE}
housing <- read_csv("Housing.csv")
housing
```

# Housing Prices 


```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(housing, aes(x=price)) +
  geom_histogram(binwidth=50000, fill="skyblue", color="darkblue") + 
  labs(title="Histogram of house prices",
       x="Sale Price",
       y="Count")
theme_minimal()
```

I can see the price is right skewed, which can make it difficult to make assumptions about.


<br>


## Explain any transformations or modifications you made to the dataset.

The price data is right-skewed  so I will transform this column I will take the log of price to create a normal(ish) distibution. 

```{r include=FALSE, warning=FALSE, message=FALSE}
housing$Log_SalePrice <- log(housing$price)
```


```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(housing, aes(x=Log_SalePrice)) +
  geom_histogram( fill="skyblue", color="darkblue") + 
  labs(title="Histogram of house prices",
       x="Sale Price",
       y="Count")
theme_minimal()
```

Using the log of price, we now have a normal distribution, but not perfectly normal



<br>

<!-- Create a linear regression model where “area” infers price (i.e., area is used to predict price). -->

```{r include=FALSE, warning=FALSE, message=FALSE}
lm_model <- lm(Log_SalePrice ~ area, data = housing)

model_summary <- summary(lm_model)
```


<!-- Get a summary of your first model and explain your results (e.g., R2, adj R2, RMSE, MSE). -->


## Linear Model Results

```{r echo=FALSE, warning=FALSE, message=FALSE}
cat('\nThe R-squared of', model_summary$r.squared, 'or 29.5% of the variation in price can be explained by area.')

cat('\n\nThe adjusted R-squared of', model_summary$adj.r.squared, ' or 29.3%, tells us that sill only 29% of the variation can be explained by\narea, even after accounting foe the number of predictors.')

cat('\n\nThe RMSE of',  sqrt(mean(lm_model$residuals^2)), 'tells us that, on average, the predictors differ from observed data.')

cat('\n\nThe MSE of', mean(lm_model$residuals^2), 'means that, on average, the squared differences between the model\'s predicted prices and\nand actual prices are close to 0.09 units squared.')
```


<br>


<!-- Get the residuals of your model (you can use ‘resid’ or ‘residuals’ functions) and plot them. What does the plot tell you about your predictions? -->

```{r echo=FALSE, warning=FALSE, message=FALSE}
# residuals from the model
resid_lm <- resid(lm_model)


# create a data frame of redisuals
resid <- data.frame(Index=seq_along(resid_lm), Residuals=resid_lm)

# plot the residuals
ggplot(resid, aes(x = Index, y = Residuals)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals Plot (ggplot2)",
       x = "Index",
       y = "Residuals") +
  theme_minimal()
```

You can clearly see a pattern, which suggest that the relationship between area and log(price) may
mot be perfectly linear. So, this model would not be appropriate for this data.


<br>


<!-- Use a qq plot to observe your residuals. Do your residuals meet the normality assumption? -->

```{r, echo=FALSE, warning=FALSE, message=FALSE}
resid_df <- data.frame(residuals = resid(lm_model))

ggplot(resid_df, aes(sample = residuals)) +
  stat_qq(color = "blue") +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot of Model Residuals",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

Almost all point fall directly on the red line, which tells us the residuals are approximately normal
This supports the use of linear regression

<br>


## Multiple Predictor Linear Regression Model

<!-- Create a linear regression model that uses multiple predictor variables to predict Sale Price (feel free to derive new predictors from existing ones). 
Explain why you think each of these variables may add explanatory value to the model. -->

```{r echo=FALSE, warning=FALSE, message=FALSE}
mlm_model <- lm(Log_SalePrice ~ area + bedrooms + bathrooms + 
                  stories + parking + airconditioning,
                data=housing)
mlm_summary <- summary(mlm_model)
```

### Multiple Predictors and Reasoning

| Category | Reasoning |
| --- | --- |
| *Area* | Most realators values homes at so many dollars per square foot. Bigger houses sell for more money. |
| *Bedrooms* | More bathrooms are more appealing to families. |
| *Bathrooms* | More bathrooms add convenience, thus value. |
| *stories * |  More stories require less land to make a larger home. |
| *parking* | Small parking areas require more management, while more parking is easier to manage |
| *airconditioning* | Depending on location, air conditioning is practically an essential |


<br>


## Multiple Regressor Summary

<!-- Get a summary of this GLM and explain your results. -->

```{r, echo=FALSE, warning=FALSE, message=FALSE}
mlm_summary
```


<br>


### Results

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Explanation of results:

cat('\nThe coefficients show the estimated effect of each predictor on log(price), holding other variables constant.\n')
cat('\nThe R-squared and adjusted R-squared values indicate how much of the variation in log(sale price) is explained by the model.\n')
cat('\nSignificant p-values (Pr(>|t|)) for predictors suggest those variables have a statistically significant association with log(sale price).\n')
cat('\nCompare these values to the single-predictor model to assess improvement.\n')
```


<br>


<!-- Get the residuals of your second model (you can use ‘resid’ or ‘residuals’ functions) and plot them. What does the plot tell you about your predictions? -->

### Second Model' Residuals

<!-- Get a summary of this GLM and explain your results. -->

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Get the residuals from the multiple linear regression model
resid_mlm <- resid(mlm_model)

# create a data frame of these residuals
resid_mlm_df <- data.frame(Index=seq_along(resid_mlm), MResiduals=resid_mlm)

# plotting the residuals
ggplot(resid_mlm_df, aes(x=Index, y=MResiduals)) + 
  geom_point(color="skyblue") +
  geom_hline(yintercept=0, color="red", linetype="dashed") + 
  labs(title="Multiple Linear Regression Residuals",
       x="Index",
       y="MResiduals") +
  theme_minimal()
```

You can clearly see a pattern, which suggest that the relationship between area and log(price) may
not be perfectly linear. Just like the residuals for the single residual. So, this model would not be appropriate for this data.




<br>



<!-- Use a qq plot to observe your residuals. Do your residuals meet the normality assumption? -->

```{r, echo=FALSE, warning=FALSE, message=FALSE}
resid_mlm_df <- data.frame(residuals = resid(mlm_model))

ggplot(resid_mlm_df, aes(sample = residuals)) +
  stat_qq(color = "purple") +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot of Residuals (Multiple Linear Regression)",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

Almost all point fall directly on the red line as they did with just area, which tells us the residuals are approximately normal.
This supports the use of linear regression


<br>



<!-- Compare the results (e.g., R2, adj R2, RMSE, MSE) between your first and second model.  -->

```{r}

# First model: simple linear regression (area only)
r2_a <- model_summary$r.squared
adjr2_a <- model_summary$adj.r.squared
rmse_a <- sqrt(mean(lm_model$residuals^2))
mse_a <- mean(lm_model$residuals^2)

# Second model: multiple linear regression
r2_m <- mlm_summary$r.squared
adjr2_m <- mlm_summary$adj.r.squared
rmse_m <- sqrt(mean(mlm_model$residuals^2))
mse_m <- mean(mlm_model$residuals^2)


cat("First Model (area only):\n",
    "R-squared:", r2_a, "\n",
    "Adjusted R-squared:", adjr2_a, "\n",
    "RMSE:", rmse_a, "\n",
    "MSE:", mse_a, "\n\n")

cat("Second Model (multiple predictors):\n",
    "R-squared:", r2_m, "\n",
    "Adjusted R-squared:", adjr2_m, "\n",
    "RMSE:", rmse_m, "\n",
    "MSE:", mse_m, "\n")


cat("\n\n\nMultiple regression model(MRM) R-squared:",  r2_m, "is greater than Area regression model (ARM) R-squared:", r2_a, "and")
cat("\nMRM Adjusted R-squared:",  adjr2_m, "is greater than ARMl R-squared:", adjr2_a, ".\n")
cat("\n\tMeaning: MRM explains more of the variation in log(price) than the ARM.")

cat("\n\n\nMultiple regression model(MRM) Root Mean Squared Error (RMSE) :",  rmse_m, "is less than Area regression model (ARM) RMSE:", rmse_a, "and")
cat("\nMRM Mean Squared Error (MSE):",  mse_m, "is less than ARMl Mean Squared Error (MSE):", mse_a, ".\n")
cat("\n\tMeaning: MRM RMSE and MSE are less indicating MRM predictions are, on average, closer to the actual values.")
```


<!-- Does your new model show an improvement over the first? To confirm a ‘significant’ improvement between the second and first model, use ANOVA to compare them. What are the results? -->

```{r}
anova_result <- anova(lm_model, mlm_model)
anova_result
```

```{r}
cat("With a p-value of", anova_result$`Pr(>F)`,"we know the Multiple regression model is a significantly better model\ncompared to the area regression model")
```

